\chapter{Software Architecture Design}

\section{System Architecture Overview}

SantaTrail is designed as an integrated UAV training platform that combines a gamified game environment, realistic drone mission planning and simulation, and AI-based adaptive training. 
The architecture consists of six main components:

\begin{itemize}
    \item Game Environment (Unity): provides the Christmas-themed gamified interface where users interact with the system, view missions, plan routes, observe simulation, and receive feedback.
    \item Mission Planning (QGroundControl): allows users to plan drone delivery routes by defining waypoints, altitude, and navigation paths on real-world terrain.
    \item Drone Simulation (ArduPilot SITL): executes the planned mission in a simulated environment and generates telemetry data such as position, velocity, altitude, and flight behavior.
    \item Communication Layer (MAVLink): exchanges mission commands and telemetry data between QGroundControl, ArduPilot SITL, and the SantaTrail system.
    \item AI Module (Python Backend): handles mission content generation (LLM), adaptive difficulty recommendation (Contextual Bandit), training progression optimization (RL), and performance feedback.
    \item Database: stores user profiles, mission data, performance history, telemetry logs, and AI training data.
\end{itemize}

These components work together through a closed feedback loop: the user plans and executes a mission then the system collects performance data then the AI module analyzes performance and generates or recommends the next mission then the user receives feedback and progresses.

\section{Software Design}

The system operates through the SantaTrail gameplay loop, where each cycle represents one training mission.

\textbf{Main Flow: Adaptive UAV Training Session:}

\begin{enumerate}
    \item \textbf{Mission Generation:} The AI module generates a mission letter using an LLM. The letter describes a child's delivery request, location context, and environmental hints. Mission metadata (terrain, difficulty, constraints) is assigned based on the user's current skill level.
    
    \item \textbf{Present Preparation:} The user selects and packs presents in the Unity game interface. Gift properties (weight, size) affect drone performance during the mission, introducing the concept of payload planning.
    
    \item \textbf{Route Planning:} The user plans a drone delivery route using QGroundControl on a real-world GIS terrain map. The user places waypoints, sets altitude, and designs a flight path while considering terrain, obstacles, restricted areas, and delivery targets.
    
    \item \textbf{Mission Execution (Simulation):} The planned mission is sent to ArduPilot SITL via MAVLink for execution. The simulation runs and generates telemetry data including drone position, velocity, altitude, trajectory, battery usage, and control inputs. The Unity environment visualizes the drone executing the delivery in real-time 3D.
    
    \item \textbf{Data Collection:} During and after the mission, the system collects performance data such as:
    \begin{itemize}
        \item Mission success or failure
        \item Route efficiency (planned vs.\ actual path)
        \item Delivery accuracy (distance to target)
        \item Safety violations (altitude, restricted area, collision)
        \item Completion time
        \item Trajectory deviation
    \end{itemize}
    
    \item \textbf{Performance Evaluation and Feedback:} The AI module processes the collected data and evaluates user performance. The system provides personalized feedback highlighting strengths and weaknesses. The user earns coins and Christmas spirit points based on performance.
    
    \item \textbf{Adaptive Next Mission:} Based on the user's performance history and current context, the Contextual Bandit model recommends a suitable difficulty level for the next mission. The LLM then generates a new mission letter matching that difficulty. Over time, Reinforcement Learning can optimize the long-term training progression policy.
    
    \item \textbf{Progression:} As the user improves, harder missions are unlocked. After completing all training missions, the user receives a certification and can enter free play mode.
\end{enumerate}

\section{AI and Data Design}

SantaTrail uses three AI approaches, each serving a different role in the training system.

\subsection{Data Sources and Collection}

The system uses two data sources:

\subsection*{A. System-Generated Data (Primary)}

Collected from actual gameplay sessions in SantaTrail:
\begin{itemize}
    \item \textbf{Mission data:} mission parameters, difficulty, terrain, constraints.
    \item \textbf{Telemetry data:} drone position, velocity, altitude, trajectory, battery, control inputs (from ArduPilot SITL via MAVLink).
    \item \textbf{Performance data:} mission success, route efficiency, delivery accuracy, safety violations, completion time.
    \item \textbf{User data:} skill level, mission history, progression, rewards earned.
\end{itemize}

Each mission produces a structured record:
\[
(\text{user\_context}, \text{mission\_action}, \text{performance\_reward})
\]

This data is used to train and update the Contextual Bandit and Reinforcement Learning (RL) models.

\subsection*{B. Helper Dataset (Reference)}

Based on existing UAV telemetry datasets (e.g., Fixed-Wing UAS Telemetry Benchmark) for:
\begin{itemize}
    \item Guiding data structure design and feature selection.
    \item Validating that the system's telemetry collection captures meaningful flight behavior features.
    \item Providing reference patterns for feature engineering.
\end{itemize}

This dataset is not used directly for model training but serves as a design reference.

\subsection*{4.3.3 Data Preprocessing}

Before being used in the AI module, collected data undergoes:
\begin{itemize}
    \item \textbf{Data Cleaning:} remove noise, incomplete records, and irrelevant attributes.
    \item \textbf{Feature Selection:} select performance-relevant features (e.g., route efficiency, altitude deviation, safety score).
    \item \textbf{Feature Engineering:} compute derived metrics:
    \begin{itemize}
        \item Path deviation (planned vs.\ actual route)
        \item Altitude stability
        \item Delivery accuracy (distance to target)
        \item Speed consistency
        \item Safety violation count
    \end{itemize}
    \item \textbf{Normalization:} scale features for consistent model input.
\end{itemize}

\subsection{Model Design}

SantaTrail uses three AI approaches, each serving a different role in the training system.

\textbf{Components of the Model:}

\subsection*{4.3.1 AI Components}

\subsubsection*{A. Large Language Model (LLM): Mission Content Generation}

The Large Language Model (LLM) is responsible for generating mission content that makes each training session unique and contextually meaningful.

The LLM generates:
\begin{itemize}
    \item Mission letters: describing the child's delivery request, personal context, and environmental hints.
    \item Mission scenario descriptions: providing narrative context for the delivery task.
    \item Learning guidance and tips: explaining mission planning concepts relevant to the current task.
\end{itemize}

\textbf{Input to the LLM:}
\begin{itemize}
    \item Mission difficulty level
    \item Terrain type and location
    \item User skill level
    \item Mission constraints (e.g., weather, obstacles, time limit)
\end{itemize}

\textbf{Output from the LLM:}
\begin{itemize}
    \item Generated mission letter text
    \item Scenario description
    \item Contextual hints for the user
\end{itemize}

\subsubsection*{B. Contextual Bandit: Adaptive Difficulty Recommendation}

The Contextual Bandit model recommends the most suitable mission difficulty for each user based on their performance history.

\textbf{Components:}
\begin{itemize}
    \item Context: user performance features extracted from mission data (e.g., average route efficiency, safety violation rate, completion rate, improvement trend).
    \item Action: candidate difficulty levels or mission types.
    \item Reward: a function based on mission success, user improvement, and engagement (e.g., not too easy, not too hard).
\end{itemize}

\textbf{Process:}
\begin{enumerate}
    \item After each mission, extract user performance features to form the context.
    \item For each candidate difficulty level, predict the expected reward.
    \item Select the difficulty with the highest expected reward.
    \item Apply an exploration strategy (e.g., epsilon-greedy) to avoid always selecting the same difficulty.
\end{enumerate}

A prediction model (e.g., Linear Regression, Random Forest, or XGBoost) is used to estimate the expected reward for each candidate action.

\subsubsection*{C. Reinforcement Learning (RL): Training Progression Optimization}

Reinforcement Learning (RL) can be explored to optimize the long-term training progression policy. While the Contextual Bandit selects the next mission based on immediate context, RL considers the full sequence of missions and optimizes for long-term learning outcomes.

\textbf{Components:}
\begin{itemize}
    \item State: user skill profile, mission history, current performance level.
    \item Action: training decisions such as difficulty adjustment, mission type selection, or feedback strategy.
    \item Reward: long-term learning improvement, certification readiness, and engagement retention.
\end{itemize}

\subsection{Evaluation}

The evaluation of the AI module covers both the prediction accuracy and the adaptive decision quality.

\textbf{LLM Evaluation:}

\begin{itemize}
    \item \textbf{Mean Absolute Error (MAE):} Measures how close the predicted reward is to the actual reward.
    
    \item \textbf{Mission quality assessment:} evaluate whether generated missions are coherent, contextually appropriate, and match the requested difficulty level.
    \item \textbf{Content diversity:} measure variation across generated mission letters to avoid repetitive content.
\end{itemize}

\textbf{Contextual Bandit Evaluation:}

\begin{itemize}
    \item \textbf{Mean Absolute Error (MAE):} Measures how close predicted reward is to actual reward.
    \item \textbf{Mean Squared Error (MSE):} Measures average squared error, penalizing large prediction errors.
    \item \textbf{Average Reward:} Average reward from selected missions, indicating decision quality.
    \item \textbf{Cumulative Reward:} Total reward over time, showing improvement in decision-making.
    \item \textbf{Regret:} Difference between obtained reward and best possible reward. Lower regret = better adaptation.
\end{itemize}

\textbf{RL Evaluation (Future):}

\begin{itemize}
    \item \textbf{Long-term user improvement:} whether users trained with RL-optimized progression show better skill development over time.
    \item \textbf{Certification readiness rate:} percentage of users who reach certification level within a target number of missions.
